---
title: "EV proteomic analysis for mitohondrial and surface proteins"
author: "Adrian Parrilla"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intoduction

This script will identify mitochondrial and surface proteins in large and small EVs using the Mitocarta, Cell Surface Protein atlas and Human CD markers databases.



```{r message = FALSE, warning=FALSE}
#Load needed libraries
library(tidyverse)
library(VennDiagram)
library(writexl)
library(readxl)
library(pheatmap)
library(RColorBrewer)
library(org.Hs.eg.db)
library(clusterProfiler)
library(ggrepel)
```

```{r}
#Load data and reference datasets
setwd("../data")

mitocarta <- read_excel("Mitocarta.xlsx")
vesiclepedia <- read_excel("Vesiclepedia.xlsx")
HPPs <- read_excel("Human plasma proteins.xlsx")
CD_proteins <- read_excel("CD proteins.xlsx")
CSPA <- read_excel("Surface Proteins.xlsx")

full_ds <- read.table("Full_dataset_V2.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)
```

```{r}
# Definig Mitocarta evidence to Literature, manual or APEX_matrix | IMS  (OPTIONAL)

mitocarta_sub <- mitocarta[grep(c("literature|manual|APEX_matrix|APEX_IMS"), mitocarta$Evidence),]
```

## Overview of proteins detected
```{r Venn HPPs CD}
venn.plot.Vesic.HPPs.CD <- venn.diagram(
  x = list(full_ds$Gene, HPPs$Gene, CD_proteins$CD, CSPA$`ENTREZ gene symbol`),
  category.names = c("EV prots detected" , "HPPs", "CD markers", "Surface proteins"),
  filename = NULL,
  fill = c("#F9F9F9","#0072B4","#C3DBFD", "lightblue"),
  alpha = 0.5,
  label.col = "black",
  cex = 1,
  fontfamily = "sans",
  cat.cex = 1.5,
  cat.fontfamily = "sans",
  #cat.pos = c(-10, 5, 180),
  #cat.dist = c(0.04, 0.04),
  euler.d=FALSE,
  scaled=FALSE
)

grid.newpage()
grid.draw(venn.plot.Vesic.HPPs.CD)

unlink(list.files()[grep(".log", list.files())])
```


## Identification of Mitochondrial proteins
```{r}
setwd("../outputs/")

# Count how many mitoprots have been detected
table(full_ds$Gene %in% mitocarta_sub$Symbol)

# Create the Venn diagram
venn.plot.mitoc <- venn.diagram(
  x = list(full_ds$Gene, mitocarta_sub$Symbol),
  category.names = c("EV prots detected" , "Mitocarta"),
  filename = NULL,
  fill = c("#0072B4","#C3DBFD"),
  alpha = 0.5,
  label.col = "black",
  cex = 3,
  fontfamily = "sans",
  cat.cex = 2,
  cat.fontfamily = "sans",
  cat.pos = c(-20, 20),
  cat.dist = c(0.04, 0.04),
  euler.d=FALSE,
  scaled=FALSE
)

unlink(list.files()[grep(".log", list.files())])  # remove the log files generated

grid.newpage()
grid.draw(venn.plot.mitoc)

```

```{r}
# Subset the results table to only include those rows that have mitoprots and select columns with normalised intensity
full_ds_mito <- full_ds %>% filter(full_ds$Gene %in% mitocarta_sub$Symbol) %>% select(Gene, Protein.Description, starts_with("norm_reporter_intensity"))

# Add information about the proteins from the Mitocarta dataset
full_ds_mito <- merge(full_ds_mito, mitocarta[,c("Symbol","Evidence","SubMitoLocalization", "MitoPathways")],by.x = "Gene", by.y = "Symbol", all.x = TRUE)

#Save the resutls in an excel file
write_xlsx(full_ds_mito, "Mitochondrial proteins detected in EVs.xlsx")
```

```{r}
# Assess the submitochondrial location of this proteins
table(full_ds_mito$SubMitoLocalization)

# Make a barplot with it
full_ds_mito %>% group_by(SubMitoLocalization) %>% summarise(Frequency = n()) %>% 
  ggplot(aes(x = SubMitoLocalization, y = Frequency, fill=SubMitoLocalization)) +
  geom_bar(stat = "identity")+
  geom_text(aes(label=Frequency), vjust= -1, fontface = "bold", size = 8)+
  xlab("")+
  ylim(c(0,80))+
  theme_classic()+
  theme(legend.position="none",axis.title.y = element_text(face = "bold", margin = margin(0,20,0,0), size = rel(1.1), color = 'black'),
        axis.title.x = element_text(hjust = 0.5, face = "bold", margin = margin(20,0,0,0), size = rel(1.1), color = 'black'),
        axis.text.x = element_text(face = "bold", size = rel(2.5)),
        plot.title = element_text(hjust = 0.5))
```

### Same analysis for LEV and SEV separately

```{r}
# make a new data frame for LEV and SEV
full_ds_mito_LEV <- full_ds_mito[,c(1, grep("LEV_",colnames(full_ds_mito)), 20)] 
full_ds_mito_LEV <- full_ds_mito_LEV %>% filter(rowSums(full_ds_mito_LEV[2:8])>= 0)

full_ds_mito_SEV <- full_ds_mito[,c(1, grep("SEV_",colnames(full_ds_mito)), 20)]
full_ds_mito_SEV <- full_ds_mito_SEV %>% filter(rowSums(full_ds_mito_SEV[2:8])>= 0)
```

```{r}
full_ds_mito_LEV %>% 
  group_by(SubMitoLocalization) %>% summarise(Frequency = n()) %>% 
  ggplot(aes(x = SubMitoLocalization, y = Frequency, fill=SubMitoLocalization)) +
  geom_bar(stat = "identity")+
  geom_text(aes(label=Frequency), vjust= -1, fontface = "bold", size = 8)+
  xlab("")+
  ylim(c(0,80))+
  theme_classic()+
  theme(legend.position="none",axis.title.y = element_text(face = "bold", margin = margin(0,20,0,0), size = rel(1.1), color = 'black'),
        axis.title.x = element_text(hjust = 0.5, face = "bold", margin = margin(20,0,0,0), size = rel(1.1), color = 'black'),
        axis.text.x = element_text(face = "bold", size = rel(2.5)),
        plot.title = element_text(hjust = 0.5))
```

```{r}
full_ds_mito_SEV %>%
  filter(Gene %in% c("ALDH2", "ALDH7A1", "CKMT1B", "GRHPR", "MPST", "PLSCR3")) %>% 
  group_by(SubMitoLocalization) %>% summarise(Frequency = n()) %>% 
  ggplot(aes(x = SubMitoLocalization, y = Frequency, fill=SubMitoLocalization)) +
  geom_bar(stat = "identity")+
  geom_text(aes(label=Frequency), vjust= -1, fontface = "bold", size = 8)+
  xlab("")+
  ylim(c(0,80))+
  theme_classic()+
  theme(legend.position="none",axis.title.y = element_text(face = "bold", margin = margin(0,20,0,0), size = rel(1.1), color = 'black'),
        axis.title.x = element_text(hjust = 0.5, face = "bold", margin = margin(20,0,0,0), size = rel(1.1), color = 'black'),
        axis.text.x = element_text(face = "bold", size = rel(2.5)),
        plot.title = element_text(hjust = 0.5))
```

```{r}
# Save proteins that are uniquely located in LEV or SEV
full_ds_mito_LEV_unique <- full_ds_mito_LEV[which(!full_ds_mito_LEV$Gene %in% full_ds_mito_SEV$Gene),]
#write_xlsx(full_ds_mito_LEV_unique, "Mitoprots LEV unique.xlsx")

full_ds_mito_SEV_unique <- full_ds_mito_SEV[which(!full_ds_mito_SEV$Gene %in% full_ds_mito_LEV$Gene),]
#write_xlsx(full_ds_mito_SEV_unique, "Mitoprots SEV unique.xlsx")
```

```{r}
# Make a heatmap with mitoprots
heatmap_data <- full_ds_mito %>% select(Gene, starts_with("norm_reporter_intensity")) %>% column_to_rownames("Gene")
# Remove the prefix from the column names
colnames(heatmap_data)<- gsub("norm_reporter_intensity_", "", colnames(heatmap_data))

pheatmap(heatmap_data,
         show_rownames = T,
         cluster_rows = F, cluster_cols = F, scale = "row",
         colorRampPalette(colors = c("blue", "white", "red"))(30),
         main = "Mitochondrial proteins in Old vs Young EVs", 
         fontsize_row = 3, fontsize_col = 10)
```

```{r}
#set NA to 0 so we can do row clustering
heatmap_data_0 <- heatmap_data 
heatmap_data_0[is.na(heatmap_data_0)] <- 0

# Add Mitochondrial location as row annotations
row_annot <- data.frame(Location = full_ds_mito$SubMitoLocalization)
rownames(row_annot) <- rownames(heatmap_data_0)

pheatmap(heatmap_data_0,
         show_rownames = T,
         cluster_rows = T, cluster_cols = F, scale = "row",
         colorRampPalette(colors = c("blue", "white", "red"))(30),
         main = "Mitochondrial proteins in Old vs Young EVs",
         breaks = c(seq(-2.3,3.5,length.out=31)),
         legend_breaks = c(-1.6,0.8,3.1),
         annotation_row = row_annot,
         legend_labels = c("Low", "Medium", "High"),
         fontsize_row = 1, fontsize_col = 10)
```

Clustering keeping NAs
```{r}
#Apply Z score normalization
heatmap_data_scaled <- as.data.frame(t(apply(heatmap_data[,c(1:16)], 1, function(x) { 
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
})))

# remove those rows whose distance matrix can not be calculated in case clustering is not working
giveNAs = which(is.na(as.matrix(dist(heatmap_data_scaled))),arr.ind=TRUE)
tab = sort(table(c(giveNAs)),decreasing=TRUE)
checkNA = sapply(1:length(tab),function(i){
  sum(is.na(as.matrix(dist(heatmap_data_scaled[-as.numeric(names(tab[1:i])),]))))
})

rmv = names(tab)[1:min(which(checkNA==0))]
heatmap_data_scaled = heatmap_data_scaled[-as.numeric(rmv),]

row_annot_scl <- data.frame(Location = full_ds_mito$SubMitoLocalization[full_ds_mito$Gene %in% rownames(heatmap_data_scaled)])
rownames(row_annot_scl) <- rownames(heatmap_data_scaled)

pheatmap(heatmap_data_scaled,
         show_rownames = T,
         cluster_rows = T, cluster_cols = F, scale = "none",
         colorRampPalette(colors = c("blue", "white", "red"))(30),
         main = "Mitochondrial proteins in Old vs Young EVs",
         breaks = c(seq(-2.3,3.5,length.out=31)),
         legend_breaks = c(-1.6,0.8,3.1),
         annotation_row = row_annot_scl,
         legend_labels = c("Low", "Medium", "High"),
         fontsize_row = 3, fontsize_col = 10)
```

```{r} 
# Larve EVs mito prots only

heatmap_data_LEV <- heatmap_data_0 %>% select(starts_with("LEV")) 
heatmap_data_LEV_0 <- heatmap_data_LEV %>% filter(rowSums(heatmap_data_LEV)>= 0)

pheatmap(heatmap_data_LEV_0,
         show_rownames = T,
         cluster_rows = F, cluster_cols = F,
         scale = "row",
         colorRampPalette(colors = c("blue", "white", "red"))(30),
         main = "Mitochondrial proteins in Old vs Young LEVs",
         #legend_breaks = c(seq(-1,3,length.out=4)),
         #legend_labels = c("NAs", "Low", "Medium", "High"),
         fontsize_row = 1, fontsize_col = 10)
```

```{r GO analysis}
#GO analysis

ego.mito.LEV.BP <- enrichGO(gene = full_ds_mito_LEV_unique$Gene,
                OrgDb         = 'org.Hs.eg.db',
                ont           = "BP", #it can be CC and MF
                keyType       = "SYMBOL", #it can also be "ENSEMBL", "ENTREZ"
                pAdjustMethod = "BH",   
                qvalueCutoff = 0.05,
                readable      = TRUE)

dotplot(ego.mito.LEV.BP, showCategory=15, font.size= 9)

ego.mito.LEV.CC <- enrichGO(gene = full_ds_mito_LEV_unique$Gene,
                OrgDb         = 'org.Hs.eg.db',
                ont           = "CC",
                keyType       = "SYMBOL",
                pAdjustMethod = "BH",   
                qvalueCutoff = 0.05,
                readable      = TRUE)

dotplot(ego.mito.LEV.CC, showCategory=15, font.size= 9)

ego.mito.SEV.BP <- enrichGO(gene = full_ds_mito_SEV_unique$Gene,
                            OrgDb         = 'org.Hs.eg.db',
                            ont           = "BP",
                            keyType       = "SYMBOL", #it can also be "ENSEMBL", "ENTREZ"
                            pAdjustMethod = "BH",   
                            qvalueCutoff = 0.05,
                            readable      = TRUE)

dotplot(ego.mito.SEV.BP, showCategory=15, font.size= 9)
```

## Surfaceome analisys: Identification of surface markers
```{r Surfaceome analisys}
# make a vector for the alternative names of CD markers
CD_alternative <- unlist(strsplit(CD_proteins$`Alternative name`, ",\\s*"))
surf <- c()

# For each gene detected, look for coincidences in the 3 datasets
for (i in full_ds$Gene){
  if (i %in% CSPA$`ENTREZ gene symbol` | i %in%  CD_proteins$CD | i %in% CD_alternative){
    surf <- append(surf, i)
  }
}

full_ds_surf <- full_ds[full_ds$Gene %in% surf,]
full_ds_surf <- full_ds_surf%>% select(Gene, Protein.Description, starts_with("norm_reporter_intensity"))
write_xlsx(full_ds_surf, "Surface proteins detected in EVs.xlsx")

```

```{r}
# Heatmap
rownames(full_ds_surf)<- NULL

heatmap_surf <- full_ds_surf %>% select(Gene, starts_with("norm_reporter_intensity")) %>% column_to_rownames("Gene")
colnames(heatmap_surf)<- gsub("norm_reporter_intensity_", "", colnames(heatmap_surf)) 

heatmap_surf_0 <- heatmap_surf
heatmap_surf_0[is.na(heatmap_surf_0)] <- 0

pheatmap(heatmap_surf_0,
         show_rownames = T,
         cluster_rows = F, cluster_cols = F,
         scale = "row",
         colorRampPalette(colors = c("blue", "white", "red"))(35),
         main = "Surface markers analysis of Old vs Young EVs", 
         fontsize_row = 1, fontsize_col = 10)
```

```{r}
pheatmap(heatmap_surf_0,
         show_rownames = T,
         cluster_rows = T, cluster_cols = F,
         scale = "row",
         colorRampPalette(colors = c("blue", "white", "red"))(35),
         main = "Surface markers analysis of Old vs Young EVs", 
         fontsize_row = 1, fontsize_col = 10)
```

```{r}
# Save unique surface proteins for LEV and SEV
full_ds_surf_LEV <- full_ds_surf[,c(1, grep("LEV_",colnames(full_ds_surf)))]
full_ds_surf_LEV <- full_ds_surf_LEV[which(rowSums(full_ds_surf_LEV[,3:9])>=0),]

full_ds_surf_SEV <- full_ds_surf[,c(1, grep("SEV_",colnames(full_ds_surf)))]
full_ds_surf_SEV <- full_ds_surf_SEV[which(rowSums(full_ds_surf_SEV[,3:9])>=0),]

full_ds_surf_LEV_unique <- full_ds_surf_LEV[which(!full_ds_surf_LEV$Gene %in% full_ds_surf_SEV$Gene),]
full_ds_surf_LEV_unique <- merge(full_ds_surf_LEV_unique, full_ds_surf[,c(1,2)])
write_xlsx(full_ds_surf_LEV_unique, "Surface prots LEV unique.xlsx")

full_ds_surf_SEV_unique <- full_ds_surf_SEV[which(!full_ds_surf_SEV$Gene %in% full_ds_surf_LEV$Gene),]
full_ds_surf_SEV_unique <- merge(full_ds_surf_SEV_unique, full_ds_surf[,c(1,2)])
write_xlsx(full_ds_surf_SEV_unique, "Surface prots SEV unique.xlsx")
```

```{r}
# Venn diagram for LEV and SEV surface proteins
venn.plot.surface.markers.LEV.SEV <- venn.diagram(
  x = list(full_ds_surf_LEV$Gene, full_ds_surf_SEV$Gene),
  category.names = c("LEV", "SEV"),
  filename = NULL,
  fill = c("#0072B4","#C3DBFD"),
  alpha = 0.5,
  label.col = "black",
  cex = 2,
  fontfamily = "sans",
  cat.cex = 2,
  cat.fontfamily = "sans",
  cat.pos = c(0, 0),
  cat.dist = c(0.06, 0.06),
  euler.d=FALSE,
  scaled=FALSE
)

grid.newpage()
grid.draw(venn.plot.surface.markers.LEV.SEV)

unlink(list.files()[grep(".log", list.files())])
```

```{r}
# GO analysis of the LEV surface proteins

LEV_surf_ego_BP <- enrichGO(gene = full_ds_surf_LEV_unique$Gene,
                            OrgDb         = 'org.Hs.eg.db',
                            ont           = "BP",
                            keyType       = "SYMBOL", #it can also be "ENSEMBL", "ENTREZ"
                            pAdjustMethod = "BH",   
                            pvalueCutoff  = 0.05,
                            readable      = TRUE)


LEV_surf_ego_CC <- enrichGO(gene = full_ds_surf_LEV_unique$Gene,
                            OrgDb         = 'org.Hs.eg.db',
                            ont           = "CC",
                            keyType       = "SYMBOL", #it can also be "ENSEMBL", "ENTREZ"
                            pAdjustMethod = "BH",   
                            pvalueCutoff  = 0.05,
                            readable      = TRUE)

LEV_surf_ego_MF <- enrichGO(gene = full_ds_surf_LEV_unique$Gene,
                            OrgDb         = 'org.Hs.eg.db',
                            ont           = "MF",
                            keyType       = "SYMBOL", #it can also be "ENSEMBL", "ENTREZ"
                            pAdjustMethod = "BH",   
                            pvalueCutoff  = 0.05,
                            readable      = TRUE)

dotplot(LEV_surf_ego_BP, showCategory=10, font.size = 10)
dotplot(LEV_surf_ego_CC, showCategory=10, font.size = 10)
dotplot(LEV_surf_ego_MF, showCategory=10, font.size = 10)
```

```{r}
# GO analysis of the SEV surface proteins

SEV_surf_ego_BP <- enrichGO(gene = full_ds_surf_SEV_unique$Gene,
                            OrgDb         = 'org.Hs.eg.db',
                            ont           = "BP",
                            keyType       = "SYMBOL", #it can also be "ENSEMBL", "ENTREZ"
                            pAdjustMethod = "BH",   
                            pvalueCutoff  = 0.05,
                            readable      = TRUE)

SEV_surf_ego_CC <- enrichGO(gene = full_ds_surf_SEV_unique$Gene,
                            OrgDb         = 'org.Hs.eg.db',
                            ont           = "CC",
                            keyType       = "SYMBOL", #it can also be "ENSEMBL", "ENTREZ"
                            pAdjustMethod = "BH",   
                            pvalueCutoff  = 0.05,
                            readable      = TRUE)

SEV_surf_ego_MF <- enrichGO(gene = full_ds_surf_SEV_unique$Gene,
                            OrgDb         = 'org.Hs.eg.db',
                            ont           = "MF",
                            keyType       = "SYMBOL", #it can also be "ENSEMBL", "ENTREZ"
                            pAdjustMethod = "BH",   
                            pvalueCutoff  = 0.05,
                            readable      = TRUE)

dotplot(SEV_surf_ego_BP, showCategory=10, font.size = 10)
dotplot(SEV_surf_ego_CC, showCategory=10, font.size = 10)
dotplot(SEV_surf_ego_MF, showCategory=10, font.size = 10)
```

